A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-07-06 doc/popl2013/intro.tex
updates

In the past decade machine architecture, particularly at the high
performance end of the spectrum, has undergone a revolution.  The
latest supercomputers are now complex hierarchies of many different
kinds of computing technologies: networks of nodes at the top level,
multiple chips per node, mutiple cores within a chip, and, most
recently, multiple accelerators (usually GPUs) per node, which 
can be further decomposed into several levels.   In this paper, we present the operational and static
semantics of Legion, a programming model targeted at providing an
appropriate level of abstraction for programming such machines, one
that is both sufficiently high-level to be portable while still
exposing aspects that are crucial to performance.

Briefly, in Legion data is organized in a hierarchy of {\em regions}
and subregions and computation is organized in a hierarchy of {\em
tasks} and subtasks operating on regions.  Regions and tasks interact
through a static system of {\em region permissions} specifying the
operations a task may perform on a region argument (read,
write, or reduce) and novel {\em region coherence} annotations that
express whether other tasks may perform operations concurrently on
the region (exclusive, atomic, or simultaneous).  We prove the
soundness of the permissions and coherence system and use these two
results to prove a third result: if two tasks $t_1$ and $t_2$ are {\em
dependent} (roughly, must be executed in program order to guarantee
the correct answer), then the unique ancestors of $t_1$ and $t_2$ in
the task hierarchy that are {\em siblings} (have the same parent task)
are also dependent.  This theorem is the basis for correct distributed
task scheduling in a Legion implementation, which is crucial for
performance; a centralized scheduler would be a bottleneck
because of the huge latencies involved in moving data and computation
in the target class of machines.

@@

Line 1 defines a {\em region relationship} {\tt rr[rll, rlr, rtl, rtr]}.  
Region relationships are bounded existential types.  The
{\tt tree} region relationship defines the type of a binary tree
node. The naming convention followed throughout the example is that
regions beginning with {\tt rl} name regions 
holding list elements and regions beginning with {\tt rt} name
regions holding tree elements.  The most important part is the {\em
  subregion} and {\em disjointness} constraints that describe the
relationships between the various regions.  Here {\tt rl} is the
region holding all list elements and {\tt rls} and {\tt rt} are the
regions holding, respetively, the subset of list elements and the tree
nodes for this tree.  Note that {\tt rls} is required to be a
subregion of {\tt rl}, which means that all the values in region {\tt
  rls} are also in region {\tt rl}.  The body of the {\tt tree} region
relationship declares the shape of a node of the tree, which includes
a pointer to a list element in {\tt rls} and pointers to two
subtrees. The region relation says that there exist four regions {\tt
  rll, rlr, rtl}, and {\tt rtr} such that the left subtree has list
and tree elements that are in subregions {\tt rll} and {\tt rtl} of
the corresponding list and tree regions {\tt rls} and {\tt rt} for the
entire tree; the right subtree's list and tree elements in subregions
{\tt rlr} and {\tt rtr} are also subregions of {\tt rls} and {\tt rt}
respectively.  Furthermore, the constraints $\tt rll \ast rlr$ and
$\tt rtl \ast rtr$ guarantee that the left and right subtrees are
disjoint.

>>>>

In the past decade machine architecture, particularly at the high
performance end of the spectrum, has undergone a revolution.  The
latest supercomputers are now complex hierarchies of many different
kinds of computing technologies: networks of nodes at the top level,
multiple chips per node, mutiple cores within a chip, and, most
recently, multiple accelerators (usually GPUs) per node, which 
can themselves be decomposed into several levels.   We present the operational and static
semantics of Legion, a programming model targeted at providing an
appropriate level of abstraction for programming such machines, one
that is both sufficiently high-level to be portable while still
exposing aspects that are crucial to performance.

Briefly, in Legion data is organized in a hierarchy of {\em regions}
and subregions and computation is organized in a hierarchy of {\em
tasks} and subtasks operating on regions.  Regions and tasks interact
through a static system of {\em region permissions} specifying 
operations a task may perform on a region argument (read,
write, or reduce) and novel {\em region coherence} annotations that
express what other tasks may do concurrently with
the region (exclusive, atomic, or simultaneous).  We prove the
soundness of the permissions and coherence system and use these two
results to prove a third result: if two tasks $t_1$ and $t_2$ are {\em
dependent} (roughly, must be executed in program order to guarantee
the correct answer), then the unique ancestors of $t_1$ and $t_2$ in
the task hierarchy that are {\em siblings} (have the same parent task)
are also dependent.  This theorem is the basis for correct distributed
task scheduling in a Legion implementation, which is crucial for
performance; a centralized scheduler would be a bottleneck
because of the huge latencies involved in moving data and computation
in the target class of machines.

@@

Line 1 defines a {\em region relationship}, which is a bounded existential type.  The
{\tt tree} region relationship defines the type of a binary tree
node. The naming convention followed throughout the example is that
regions beginning with {\tt rl} name regions 
holding list elements and regions beginning with {\tt rt} name
regions holding tree elements.  The most important part is the {\em
  subregion} and {\em disjointness} constraints that describe the
relationships between the various regions.  Here {\tt rl} is the
region holding all list elements and {\tt rls} and {\tt rt} are the
regions holding, respetively, the subset of list elements and the tree
nodes for this tree.  Note that {\tt rls} is required to be a
subregion of {\tt rl}, which means that all the values in region {\tt
  rls} are also in region {\tt rl}.  The body of the {\tt tree} region
relationship declares the shape of a node of the tree, which includes
a pointer to a list element in {\tt rls} and pointers to two
subtrees. The region relation says that there exist four regions {\tt
  rll, rlr, rtl}, and {\tt rtr} such that the left subtree has list
and tree elements that are in subregions {\tt rll} and {\tt rtl} of
the corresponding list and tree regions {\tt rls} and {\tt rt} for the
entire tree; the right subtree's list and tree elements in subregions
{\tt rlr} and {\tt rtr} are also subregions of {\tt rls} and {\tt rt}
respectively.  Furthermore, the constraints $\tt rll \ast rlr$ and
$\tt rtl \ast rtr$ guarantee that the left and right subtrees are
disjoint.

### 2012-07-06 doc/popl2013/langdef.tex
updates

Figure~\ref{fig:langdef} defines the core subset of Legion.  Types
include booleans, integers, tuples, and pointers annotated with a list
of regions into which they may point.  There is a special type for
{\em colorings}, which are used to specify partitions of regions into
subregions.  Functions/tasks in Legion are named, accept one or more
arguments and return a value of some type.  A function also declares
what region access privileges and coherence it requires.  Function
types are universally quantified over all logical region bindings,
with all argument and result types, as well as the privileges and
coherence requirements being expressed relative to the quantified
region names.  The final additional value type that Legion uses is a
region relation, which stores a value and one or more logical regions
satisfying associated region constraints.  An instance of a region
relation can be read out of the heap and unpacked, giving new local
names to the logical regions contained in the instance.
% and tracking
%the constraints between those new logical regions.

@@

Core Legion is explicitly typed using judgments of the form
\begin{center}
$\typeenv{e}{T}$
\end{center}
where
\begin{itemize}
\item $\Gamma$ is the type environment, mapping the names of local variables and global functions
to their types,
\item $\Phi$ is the region access privileges under which the expression will be evaluated, and
\item $\Omega$ is the set of constraints that must hold between regions
\end{itemize}
As regions are hierarchical, privileges and/or constraints on a region often imply similar
privileges or constraints on subregions.  Figure~\ref{fig:closure} defines closure
operations for $\Phi^*$ and $\Omega^*$ that add implied constraints.
The type checking rules are in Figure~\ref{fig:types}.  Many of the rules are 
straightforward, but several deserve more explanation.

Core Legion has no subtyping, even for pointers into regions that are part of a 
known region hierarchy.  The region ``scope'' of a pointer can only be changed via explicit
{\em upregion} and {\em downregion} operations, which check that the ``from'' and ``to'' regions
are in the right relationship.  It would have been possible to safely infer {\em upregion} calls,
but {\em downregion} calls must be accompanied by a run-time check that the pointer really does
point into the subregion(s), so {\em upregion} was made explicit as well for simplicity.

Accesses to heap memory are performed via the {\em read}, {\em write}, and {\em reduce}
operations, which are only legal if the region(s) in a pointers type are ones for which the
enclosing function has privileges to access.

Similarly, a call to another function is only permitted if the privileges required by the 
callee are a subset of those held by the caller.  There is no way to create privileges within
a Legion function - it is limited to those given to it by its caller.

The {\em partition} operation uses a {\em coloring} to split a region into one or more
disjoint subregions.  The subregions are given names within the body of the {\em partition},
as well as constraints that assert that each subregion is contained by the parent region and
the pairwise disjointness of the subregions.  These constraints can be used locally, or packed
into region relation instances that can be unpacked and used later.

The {\em pack} expression packs together a value and the corresponding set of logical regions,
as long as the constraints in the region relationship type can be shown to hold.  The
corresponding {\em unpack} expression gives (new) names to those regions.  Neither the 
{\em partition} nor the {\em unpack} expression allows aliasing of region names, simplifying the
handling of existing variables and region names.

@@

\begin{figure}
\centering{
$\begin{array}{l}
\Omega \subseteq \Omega^* \\
r_i \leq r_j \in \Omega^*  \Rightarrow r_i \leq r_i \in \Omega^*\wedge r_j \leq r_j \in \Omega^* \\
r_i \leq r_j \in \Omega^* \wedge r_j \leq r_k \in \Omega^* \Rightarrow r_i \leq r_k \in \Omega^* \\
r_i \leq r_j \in \Omega^* \wedge r_j * r_k \in \Omega^* \Rightarrow r_i * r_k \in \Omega^* \\
r_i * r_j \in \Omega^* \Rightarrow r_j * r_i \in \Omega^* \\
\\
\Phi \subseteq \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge reads(r_j) \in \Phi^* \Rightarrow reads(r_i) \in \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge writes(r_j) \in \Phi^* \Rightarrow writes(r_i) \in \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge reduces_{id}(r_j) \in \Phi^* \Rightarrow reduces_{id}(r_i) \in \Phi^* \\
reads(r) \in \Phi^* \wedge writes(r) \in \Phi^* \wedge \\
\hspace{1cm} \Gamma(id) = \forall \otok{k}{r}. (\oton{T}), \Phi', Q' \rightarrow T_r \Rightarrow reduces_{id}(r) \in \Phi^*
\end{array}$
}
\caption{Privilege and Constraint Closure}
\label{fig:closure}
\end{figure}

@@

Operational semantics rules have the following form:

>>>>

Figure~\ref{fig:langdef} defines the core subset of Legion.  Types
include booleans, integers, tuples, and pointers, which are annotated with a list
of regions into which they may point.  There is a special type for
{\em colorings}, which are used to specify partitions of regions into
subregions.  Functions/tasks in Legion are named, accept one or more
arguments and return a value of some type.  A function also declares
the region access privileges and coherence it requires.  Function
types are universally quantified over all logical region bindings,
with all argument and result types, as well as the privileges and
coherence requirements being expressed relative to the quantified
region names.  The final additional value type that Legion uses is a
region relation, which stores a value and one or more logical regions
satisfying associated region constraints.  An instance of a region
relation can be read out of the heap and unpacked, giving new local
names to the logical regions contained in the instance.
% and tracking
%the constraints between those new logical regions.

@@

Core Legion is explicitly typed using judgments of the form
\begin{center}
$\typeenv{e}{T}$
\end{center}
where
\begin{itemize}
\item $\Gamma$ is the type environment, mapping the names of local variables and global functions
to their types,
\item $\Phi$ is the region access privileges under which the expression is evaluated, and
\item $\Omega$ is the set of constraints that must hold between regions
\end{itemize}
As regions are hierarchical, privileges and/or constraints on a region often imply similar
privileges or constraints on subregions.  Figure~\ref{fig:closure} defines closure
operations for $\Phi^*$ and $\Omega^*$ that add implied constraints.
The type checking rules are in Figure~\ref{fig:types}.  Many of the rules are 
straightforward, but several deserve more explanation.

Core Legion has no subtyping, but a pointer's region can be changed via explicit
{\em upregion} and {\em downregion} operations.  {\em Upregion} calls can be inferred,
but {\em downregion} must be accompanied by a run-time check that the pointer really does
point into the subregion(s), so {\em upregion} is made explicit as well for simplicity.

Accesses to heap memory are performed via the {\em read}, {\em write}, and {\em reduce}
operations, which are only legal if the regions in a pointer's type are ones for which the
enclosing function has privileges.
Similarly, a call to another function is permitted only if the privileges required by the 
callee are a subset of those held by the caller.  There is no way to create privileges within
a Legion function - it is limited to privileges given to it by its caller.

The {\em partition} operation uses a {\em coloring} to split a region into 
disjoint subregions.  The subregions are given names within the body of the partition,
as well as constraints asserting subregions are pairwise disjoint and contained within the parent region.
%These constraints can be used locally, or packed into region relation instances that can be unpacked and used later.

The {\em pack} expression packs together a value and the corresponding logical regions,
provided the region relation's constraints hold.  The
corresponding {\em unpack} expression gives (new) names to those regions.  For simplicitly, neither the 
{\em partition} nor the {\em unpack} expression allows shadowing of region names.

@@

\begin{figure}
\centering{
$\begin{array}{l}
\Omega \subseteq \Omega^* \\
r_i \leq r_j \in \Omega^*  \Rightarrow r_i \leq r_i \in \Omega^*\wedge r_j \leq r_j \in \Omega^* \\
r_i \leq r_j \in \Omega^* \wedge r_j \leq r_k \in \Omega^* \Rightarrow r_i \leq r_k \in \Omega^* \\
r_i \leq r_j \in \Omega^* \wedge r_j * r_k \in \Omega^* \Rightarrow r_i * r_k \in \Omega^* \\
r_i * r_j \in \Omega^* \Rightarrow r_j * r_i \in \Omega^* \\
\\
\Phi \subseteq \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge reads(r_j) \in \Phi^* \Rightarrow reads(r_i) \in \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge writes(r_j) \in \Phi^* \Rightarrow writes(r_i) \in \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge reduces_{id}(r_j) \in \Phi^* \Rightarrow reduces_{id}(r_i) \in \Phi^* \\
reads(r) \in \Phi^* \wedge writes(r) \in \Phi^* \wedge id \in \mbox{\em functions}(\Gamma) \Rightarrow reduces_{id}(r) \in \Phi^*
\end{array}$
}
\caption{Privilege and Constraint Closure}
\label{fig:closure}
\end{figure}

@@

Core Legion's operational semantics rules have the following form:
